---
title: "Stan Introduction"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Need for Stan
Part of the difficulty of Bayesian statistics for us is the R is not equipped to handle some of the necessary procedures. For example, sampling from a posterior distribution is an essential part of Bayesian inference, but if we are using MCMC or a variation of it as part of our sampling (and we likely are) writing the code in R can be long and comlicated. Even running the code neccessary may take an impractical amount of time.

Stan is a programming language seperate from R designed to help in Bayesian analysis. Stan removes the dozens of lines of code that we would have needed to write, and instead asks for us to fill in three basics blocks of information: 'data', 'parameters', and 'model'.

We'll explain more about each of these components and show how to use Stan within Rstudio with the example code below.


```{r}
# Run all R code in a new R script file called RStan_introduction
# The package below will install Stan to Rstudio, and you will be able to create Stan files.
install.packages("rstan", repos = "https://cloud.r-project.org/", dependencies = TRUE)
```



```{r}
pkgbuild::has_build_tools(debug = TRUE) #This line should return output 'TRUE'

```



```{r}
library(rstan)
#If this is your first Rstan installation, restart R. If the installation has not gone correctly, a more in depth guide can be found at https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started
```

Now we should have rstan installed. Restart R! As with most new packages, this should get everything we need working.

Go to the file tab in the top left corner of Rstudio. Select new file. You should now see 'Stan file' in this list. Select it, and return here.

If Stan file is not an option, something went wrong in setup. Return to this website https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started for more help.

If the Stan file is now open, perfect, save it as my_stan_model and leave it there for now. We need to make some fake data for our Stan model.

##Create our data

```{r}
mymodeldata = list(
N = 100,
Y = rnorm(N, 1.6, 0.2)
)
hist(Y)
```

So we have some very simple data that will work for our stan model. We have 100 data points, the data is normally distributed around a mean of 1.6, and has a standard deviation of 0.2. From here we will start building stan in your new stan file.

Clear any default text out of your Stan file and paste the following.


##Build Stan model


Quick note, Stan files use double slash marks for comments // and lines within blocks should end with a semi colon ;.


data {

}

parameters {

}

model {

}


These are the blocks I referred to earlier, and are the most used. There are others you'll learn about later.
Lets start with the first, data. Pretty straitforward, we need to tell the model what data we're working with. Down below you'll see how we format it.


data {
  int N; 
  real Y[N];
}

parameters {

}

model {

}


First we include our sample size. int because we are dealing with integers, a discrete count. Second, our observations, Y[N]. Real because our data are continuous. N is in brackets to tell stan that we have an array of data points for N individuals (100 in this case).


data {
  int N; 
  real Y[N];
}

parameters {
  real mu;
  real<lower=0> sigma;
}

model {

}


Our model section includes the parameters we are trying to estimate from our data. In our simple model here, we only are worried about mu and sigma. Both are continuous, so type real, but we want to make sure the model knows that sigma cannot be negative. Thus we add in the <lower=0> modifier. That's it for parameters.


data {
  int N; 
  real Y[N];
}

parameters {
  real mu;
  real<lower=0> sigma;
}

model {
  for(i in 1:N)
    Y[i] ~ normal(mu, sigma);
  mu ~ normal(1.7, 0.3)
  sigma ~ cauchy(0, 1);
}


The model block gets a little more complicated, as will parameters and data in more complicated models that we will see later. For right now, this will do. We start by defining our likelyhood here, in which we make a for loop that will run all points of our data Y, assuming a normal distribution. Then we define our priors for both our parameters, mu and sigma. In this code, we assume a prior very close to our data, but tweek the numbers here to see how priors interact with the likelyhood.


##Using our Stan model in R
Lets return to our Rscript.

Since we already have the rstan library loaded, we can run our data through the stan model and look at some of the important results.

```{r}
fit = stan(
  file = here::here("my_stan_model"),
  data = mymodeldata,
  seed = 42
)
```




